First steps towards a formalization of Forcing \\ Revisions made to the paper

### 2018-11-19 LSFA/cover_letter.tex
quedó un "his" de otra versión de esta carta.
Miguel haciendo sus cosas tarde

In first place we want to thank the referees for reading the paper very
carefully and for his detailed and constructive criticism. Below we
detail the main changes that have been made%; also, a diff-PDF is
%provided that shows all the changes in color
.

@@

The section on ``Names'' includes a thorough discussion on recursive
definitions in Isabelle/ZF and two new applications. One is the
proof that the generic
extension $M[G]$ is a transitive set. The second one is even more basic  and
was missing in the two previous versions of the manuscript: The
generic filter $G$ belongs to $M[G]$.

We take this opportunity to answer a question by Reviewer 3 from the
first round of corrections. Obtaining the results of this paper per se
was not very difficult. But as this Reviewer points out, the setting
up for the project, planning a strategy, and even studying Paulson's
libraries%
%%%%%%%%%%%%%%%%%%%%
\footnote{Even today we find ourselves inadvertently reproving some
  variant of a lemma in Paulson's development on set theory.}
%%%%%%%%%%%%%%%%%%%%
took almost an year of work. With the hindsight of this planning period,
we would like to point out that most of the results stated in the
conclusion section involve some serious development; even  a
publication for each item. It is therefore not feasible to include
none of them in this revision.
%
%
\bibliographystyle{arXiv/mi-estilo-else}
\bibliography{citados}
\end{document}
%%% Local Variables: 
%%% mode: latex
%%% ispell-local-dictionary: "american"
%%% End:

>>>>

In first place we want to thank the referees for reading the paper very
carefully and for their detailed and constructive criticism. Below we
detail the main changes that have been made%; also, a diff-PDF is
%provided that shows all the changes in color
.

@@

The section on ``Names'' includes a thorough discussion on recursive
definitions in Isabelle/ZF and two new applications. One is the proof
that the generic extension $M[G]$ is a transitive set and that the
generic filter $G$ belongs to $M[G]$.

We take this opportunity to answer a question by Reviewer 3 from the
first round of corrections. Obtaining the results of this paper per se
was not very difficult. But as this Reviewer points out, the setting
up for the project, planning a strategy, and even studying Paulson's
libraries%
%%%%%%%%%%%%%%%%%%%%
\footnote{Even today we find ourselves inadvertently reproving some
  variant of a lemma in Paulson's development on set theory.}
%%%%%%%%%%%%%%%%%%%%
took almost an year of work. With the hindsight of this planning
period, we would like to point out that most of the results stated in
the conclusion section involve some serious development; and it might
even result in a publication for each item. It is therefore not
feasible to include any of them in this revision.
%
%
\bibliographystyle{arXiv/mi-estilo-else}
\bibliography{citados}
\end{document}
%%% Local Variables: 
%%% mode: latex
%%% ispell-local-dictionary: "american"
%%% End:

### 2018-11-19 LSFA/names.tex
Miguel haciendo sus cosas tarde

We can finally define the generic extension of $M$ by $G$, also
setting up the notation $M[G]$ for it:
\begin{isabelle}
\isacommand{definition}\isamarkupfalse%
\isanewline
\ \ GenExt\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}i{\isachardoublequoteclose}\ \ \ \ \ {\isacharparenleft}{\isachardoublequoteopen}M{\isacharbrackleft}{\isacharunderscore}{\isacharbrackright}{\isachardoublequoteclose}{\isacharparenright}\ \isakeyword{where} \isanewline
\ \ {\isachardoublequoteopen}GenExt{\isacharparenleft}G{\isacharparenright}{\isacharequal}{\isacharequal}\ {\isacharbraceleft}val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}{\isachardot}\ {\isasymtau}\ {\isasymin}\ M{\isacharbraceright}{\isachardoublequoteclose}
\end{isabelle}

@@

\begin{isabelle}
\isacommand{definition}\isamarkupfalse%
\ \isanewline
\ \ Hcheck\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isacharbrackleft}i{\isacharcomma}i{\isacharbrackright}\ {\isasymRightarrow}\ i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
\ \ {\isachardoublequoteopen}Hcheck{\isacharparenleft}z{\isacharcomma}f{\isacharparenright}\ \ {\isacharequal}{\isacharequal}\ {\isacharbraceleft}\ {\isacharless}f{\isacharbackquote}y{\isacharcomma}one{\isachargreater}\ {\isachardot}\ y\ {\isasymin}\ z{\isacharbraceright}{\isachardoublequoteclose}\isanewline
\isanewline
\isacommand{definition}\isamarkupfalse%
\isanewline
\ \ check\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i\ {\isasymRightarrow}\ i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
\ \ {\isachardoublequoteopen}check{\isacharparenleft}x{\isacharparenright}\ {\isacharequal}{\isacharequal}\ wfrec{\isacharparenleft}Memrel{\isacharparenleft}eclose{\isacharparenleft}{\isacharbraceleft}x{\isacharbraceright}{\isacharparenright}{\isacharparenright}{\isacharcomma}\ x\ {\isacharcomma}\ Hcheck{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}
Here, \isa{eclose} returns the (downward) $\in$-closure of its
argument. The main result is stated as follows:

\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ valcheck\ {\isacharcolon}\ {\isachardoublequoteopen}y\ {\isasymin}\ M\ {\isasymLongrightarrow}\ one\ {\isasymin}\ G\ {\isasymLongrightarrow}\ {\isasymforall}x{\isasymin}M{\isachardot}\ check{\isacharparenleft}x{\isacharparenright}\ {\isasymin}\ M\ {\isasymLongrightarrow}\ val{\isacharparenleft}G{\isacharcomma}check{\isacharparenleft}y{\isacharparenright}{\isacharparenright}\ \ {\isacharequal}\ y{\isachardoublequoteclose}
\end{isabelle}

On the one hand, the only requirement on the set \isa{G} is that it
contains the top element of the poset \isa{P}. On the other hand, it
is necessary that all the relevant names are indeed in $M$ (i.e.,
$\chk(x) \in M$ if $x\in M$). It requires a serious development to
fulfill this assumption. One of the hardest parts of Paulson's
formalization of constructibility involves showing that models are
closed under recursive construction. We will eventually formalize that
if $M\models\ZFC$ and the arguments of \isa{wfrec} are in $M$, then
its value also is. This will require to adapt to ctm models several
locales defined in \cite{paulson_2003} that were intended to be used
for the class of constructible sets.

>>>>

We can finally define the generic extension of $M$ by $G$, also
setting up the notation $M[G]$ for it:
\begin{isabelle}
\isacommand{definition}\isamarkupfalse%
\isanewline
\ \ GenExt\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isasymRightarrow}i{\isachardoublequoteclose}\ \ \ \ \ {\isacharparenleft}{\isachardoublequoteopen}M{\isacharbrackleft}{\isacharunderscore}{\isacharbrackright}{\isachardoublequoteclose}{\isacharparenright}\ \isakeyword{where} \isanewline
\ \ {\isachardoublequoteopen}GenExt{\isacharparenleft}G{\isacharparenright}{\isacharequal}{\isacharequal}\ {\isacharbraceleft}val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}{\isachardot}\ {\isasymtau}\ {\isasymin}\ M{\isacharbraceright}{\isachardoublequoteclose}
\end{isabelle}
It is conventional in Isabelle/ZF to define introduction and
destruction rules for definitions like \isa{GenExt}; in our case, it is
enough to know $x\in M$ in order to know $\val(G,x) \in M[G]$:
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ GenExtI{\isacharcolon}\ \ {\isachardoublequoteopen}x\ {\isasymin}\ M\ {\isasymLongrightarrow}\ val{\isacharparenleft}G{\isacharcomma}x{\isacharparenright}\ {\isasymin}\ M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}
\end{isabelle}
\noindent The destruction rule corresponding to the generic extension says
that any $x \in M[G]$ comes from some $\tau \in M$ via $\val$. 
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ GenExtD{\isacharcolon}\ \ {\isachardoublequoteopen}x\ {\isasymin}\ M{\isacharbrackleft}G{\isacharbrackright}\ {\isasymLongrightarrow}\ {\isasymexists}{\isasymtau}{\isasymin}M{\isachardot}\ x\ {\isacharequal}\ val{\isacharparenleft}G{\isacharcomma}{\isasymtau}{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}

@@

\begin{isabelle}
\isacommand{definition}\isamarkupfalse%
\ \isanewline
\ \ Hcheck\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isacharbrackleft}i{\isacharcomma}i{\isacharbrackright}\ {\isasymRightarrow}\ i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
\ \ {\isachardoublequoteopen}Hcheck{\isacharparenleft}z{\isacharcomma}f{\isacharparenright}\ \ {\isacharequal}{\isacharequal}\ {\isacharbraceleft}\ {\isacharless}f{\isacharbackquote}y{\isacharcomma}one{\isachargreater}\ {\isachardot}\ y\ {\isasymin}\ z{\isacharbraceright}{\isachardoublequoteclose}\isanewline
\isanewline
\isacommand{definition}\isamarkupfalse%
\isanewline
\ \ check\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i\ {\isasymRightarrow}\ i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
\ \ {\isachardoublequoteopen}check{\isacharparenleft}x{\isacharparenright}\ {\isacharequal}{\isacharequal}\ wfrec{\isacharparenleft}Memrel{\isacharparenleft}eclose{\isacharparenleft}{\isacharbraceleft}x{\isacharbraceright}{\isacharparenright}{\isacharparenright}{\isacharcomma}\ x\ {\isacharcomma}\ Hcheck{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}
Here, \isa{eclose} returns the (downward) $\in$-closure of its
argument. The main lemmas about $\val$ and $\chk$ require some
instances of replacement for $M$; we set up a locale to assemble these
assumptions:
\begin{isabelle}
\isacommand{locale}\isamarkupfalse%
\ M{\isacharunderscore}extra{\isacharunderscore}assms\ {\isacharequal}\ forcing{\isacharunderscore}data\ {\isacharplus}\isanewline
\ \ \isakeyword{assumes}\ check{\isacharunderscore}in{\isacharunderscore}M\ {\isacharcolon}\ \ \ \ \ \  {\isachardoublequoteopen}{\isasymAnd}x{\isachardot}\ x\ {\isasymin}\ M\ {\isasymLongrightarrow}\ check{\isacharparenleft}x{\isacharparenright}\ {\isasymin}\ M{\isachardoublequoteclose}\isanewline
\ \ \ \ \ \ \isakeyword{and}\ sats{\isacharunderscore}upair{\isacharunderscore}ax{\isacharcolon}\ \ \ \ {\isachardoublequoteopen}upair{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \ \ \ \ \isakeyword{and}\ repl{\isacharunderscore}check{\isacharunderscore}pair\ {\isacharcolon}\ {\isachardoublequoteopen}strong{\isacharunderscore}replacement{\isacharparenleft}{\isacharhash}{\isacharhash}M{\isacharcomma}{\isasymlambda}p\ y{\isachardot}\ y\ {\isacharequal}{\isacharless}check{\isacharparenleft}p{\isacharparenright}{\isacharcomma}p{\isachargreater}{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}

The first assumption asserts that all the relevant names are indeed in
$M$ (i.e., $\chk(x) \in M$ if $x\in M$) and it is needed to prove that
$\val(G,\chk(x))=x$. It will take a serious effort to fulfill this
assumption: One of the hardest parts of Paulson's formalization of
constructibility involves showing that models are closed under
recursive construction. We will eventually formalize that if
$M\models\ZFC$ and the arguments of \isa{wfrec} are in $M$, then its
value also is. This will require to adapt to ctm models several
locales defined in \cite{paulson_2003} that were intended to be used
for the class of constructible sets. Notice that the only requirement
on the set \isa{G} is that it contains the top element of the poset
\isa{P}.

\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ valcheck\ {\isacharcolon}\ \isanewline
\ \ \isakeyword{assumes}\ {\isachardoublequoteopen}one\ {\isasymin}\ G{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\ {\isachardoublequoteopen}y\ {\isasymin}\ M\ {\isasymLongrightarrow}\ val{\isacharparenleft}G{\isacharcomma}check{\isacharparenleft}y{\isacharparenright}{\isacharparenright}\ \ {\isacharequal}\ y{\isachardoublequoteclose}
\end{isabelle}

\subsection{Basic results about the generic extension}

We turn now to prove that $M[G]$ is transitive and $G\in
M[G]$. Showing that $M[G]$ is transitive amounts to prove $y \in M[G]$
for any $x \in M[G]$ and $y \in x$. 
% \begin{isabelle}
% \isacommand{definition}\isamarkupfalse%
% \isanewline
% \ \ Transset\ \ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isacharequal}{\isachargreater}o{\isachardoublequoteclose}\ \ \isakeyword{where}\isanewline
% \ \ \ \ {\isachardoublequoteopen}Transset{\isacharparenleft}i{\isacharparenright}\ {\isacharequal}{\isacharequal}\ {\isasymforall}x{\isasymin}i{\isachardot}\ x{\isacharless}{\isacharequal}i{\isachardoublequoteclose}
% \end{isabelle}

\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ trans{\isacharunderscore}Gen{\isacharunderscore}Ext{\isacharprime}\ {\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\ \ {\isachardoublequoteopen}x\ {\isasymin}\ M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}\ \ \isakeyword{and}\ \ {\isachardoublequoteopen}y\ {\isasymin}\ x{\isachardoublequoteclose}\ \isanewline
\ \ \isakeyword{shows}\ \ \ \ {\isachardoublequoteopen}y\ {\isasymin}\ M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}
\end{isabelle}
The proof of this lemma is straightforward because from $x \in M[G]$
we can obtain $\tau \in M$ such that $x = \val(G,\tau)$. Notice also
that using the characterization of $\val$ given by
\isa{def{\isacharunderscore}val} we can extract some
$\theta \in \dom(\tau)$ such that $y =\val(G,\theta)$; to conclude
$\val(G,\theta) \in M[G]$ it is enough to prove $\theta \in M$, which
follows from the transitivity of $M$.

In contrast, the proof that $G\in M[G]$ is more demanding. In fact, we
set $\dot{G} = \{ \langle \check{p},p\rangle \,|\, p \in P \}$ as a
putative name for $G$. Proving that $\dot{G}$ is in fact a name for
$G$ requires to prove that $\dot{G} \in M$, using an instance of
replacement for $M$ (\noindent namely that given by the assumption
\isa{repl{\isacharunderscore}check{\isacharunderscore}pair}), and
then proving that $\val(G,\dot{G})=G$.

\begin{isabelle}
\isacommand{definition}\isamarkupfalse%
\isanewline
\ \ G{\isacharunderscore}dot\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
\ \ {\isachardoublequoteopen}G{\isacharunderscore}dot\ {\isacharequal}{\isacharequal}\ {\isacharbraceleft}{\isacharless}check{\isacharparenleft}p{\isacharparenright}{\isacharcomma}p{\isachargreater}\ {\isachardot}\ p{\isasymin}P{\isacharbraceright}{\isachardoublequoteclose}
\end{isabelle}

\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ G{\isacharunderscore}dot{\isacharunderscore}in{\isacharunderscore}M\ {\isacharcolon} {\isachardoublequoteopen}G{\isacharunderscore}dot\ {\isasymin}\ M{\isachardoublequoteclose}
\isanewline\isanewline
\isacommand{lemma}\isamarkupfalse%
\ val{\isacharunderscore}G{\isacharunderscore}dot\ {\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\ {\isachardoublequoteopen}G\ {\isasymsubseteq}\ P{\isachardoublequoteclose}\ \ \isakeyword{and}\ \ 
{\isachardoublequoteopen}one\ {\isasymin}\ G{\isachardoublequoteclose}\ \isanewline
\ \ \isakeyword{shows}\ {\isachardoublequoteopen}val{\isacharparenleft}G{\isacharcomma}G{\isacharunderscore}dot{\isacharparenright}\ {\isacharequal}\ G{\isachardoublequoteclose}
\end{isabelle}

### 2018-11-19 LSFA/related.tex
Miguel haciendo sus cosas tarde

\subsection{Related work}\label{sec:related-work}
Formalization of mathematics serves many purposes. The most obvious
one is to increase reliability in a result and/or its proof. This has
been the original motivation that lead Voevodsky to gather many
researchers around \emph{homotopy type theory} and its formalization in Coq
\cite{hottbook}; the same applies to the four colour theorem (checked by
Gonthier \cite{MR2463991}) and the formidable \emph{Flyspeck} project
\cite{MR3659768} by the team conducted by Hales.

>>>>

\subsection{Related work}\label{sec:related-work}
Formalization of mathematics serves many purposes
\cite{simpson-theorem-proving-math}. The most obvious one is to
increase reliability in a result and/or its proof. This has been the
original motivation that lead Voevodsky to gather many researchers
around \emph{homotopy type theory} and its formalization in Coq
\cite{hottbook}; the same applies to the four colour theorem (checked
by Gonthier \cite{MR2463991}) and the formidable \emph{Flyspeck}
project \cite{MR3659768} by the team conducted by Hales.
